---
title: "Zero Count Filtration Step"
author: "memrekus"
date: "2023-04-04"
output: html_document
---

# Get the list of TCGA projects

```{r}

mir_projects <- gsub("_miRisoform.rds", "", list.files("miRNAseq_RDS/"))

rna_projects <- gsub("_rnaseq.rds", "", list.files("RNAseq_RDS/"))

# Ensure match
sum(rna_projects != mir_projects)

```

# Prepare TCGA data

```{r}

library(TCGAbiolinks)
library(SummarizedExperiment)
library(edgeR)
library(SummarizedExperiment)

library(data.table)
library(dplyr)


for(i in rna_projects) {  # Some projects will fail. Can execute the loop in chunks or create if statements as below
    
    message(paste("\n\n\n---------- Processing", i, "----------"))
    
    
    message("Preparing sample level metadata")
    
    rna <- readRDS(paste0("RNAseq_RDS/", i, "_rnaseq.rds"))
    
    # Prepare sample level data
    coldat <- as.data.table(colData(rna))
    
    coldat <- coldat[, colMeans(is.na(coldat)) != 1, with=F]
    
    coldat <- coldat[, !grepl("paper_", colnames(coldat)), with=F]
    
    if(!"vital_status" %in% colnames(coldat)) stop("No column name called vital_status")
    if(!"days_to_death" %in% colnames(coldat)) stop("No column name called days_to_death")
    if(!"days_to_last_follow_up" %in% colnames(coldat)) stop("No column name called days_to_last_follow_up")
    
    coldat$vital_status <- ifelse(coldat$vital_status == "Dead", 1, 0)
    coldat$days_to_event <- ifelse(coldat$vital_status == 1, coldat$days_to_death, coldat$days_to_last_follow_up)
    
    coldat$age_at_diagnosis <- coldat$age_at_diagnosis / 365
    
    colnames(coldat) <- paste0("meta.", colnames(coldat))
    
   if(!i %in% c("DLBC", "LAML", "MESO", "OV", "TGCT", "THYM")) { subtype <- TCGAquery_subtype(i)
   
   subtype <- subtype[, !colnames(subtype) %in% colnames(coldat)]
    
    colnames(subtype) <- paste0("meta.", colnames(subtype))
    
    # Remove all NA columns
    subtype <- subtype[, colMeans(is.na(subtype)) != 1]
    
    if(i == "SKCM") {colnames(subtype) <- gsub("meta.RNASEQ.CLUSTER_CONSENHIER", 
                                              "meta.RNAseq.classification", colnames(subtype))
    
 subtype$meta.RNAseq.classification <- factor(subtype$meta.RNAseq.classification, 
                                                   levels=c("immune", "keratin", "MITF-low"))
    
 # Use only primary tumor data when duplicates are present
  nonprim <- duplicated(subtype$meta.patient) & subtype$meta.ALL_PRIMARY_VS_METASTATIC != "All_Primaries"
  
  subtype <- subtype[!nonprim, ]
 
 }
    
    if(i == "PCPG") subtype$meta.patient <- substr(subtype$`meta.Sample ID`, 1, 12)
    
    subtype <- as.data.table(subtype)
    
   
    
    
    
   }
    
    
   
    
      message("Preparing RNAseq data")
    
    # Prepare RNAseq data
    rnadat <- assay(rna)
    
    rowdat <- as.data.frame(rowData(rna))
    
    rnadat <- cpm(rnadat, log=F, prior.count = 0)
    
    rnadat <- log(rnadat+1, base = 10)
    
    rnadat <- data.table(rnadat)
    
    rna_barcodes <- colnames(rnadat)
    
    # rnadat$genename <- rowdat$external_gene_name
    
    # rnadat$genename <- rowdat$gene_name
    
    # rnadat <- rnadat[, lapply(.SD, mean), by="genename"]
    
    rnadat<-as.data.frame(t(rnadat))
    
    rowdat <- rowdat %>%
  mutate(duplicated_name = duplicated(gene_name) | duplicated(gene_name, fromLast=T)) %>%
  mutate(new_name = case_when(
    duplicated_name == TRUE ~ paste(gene_name, gene_id, sep = "_"),
    duplicated_name == FALSE ~ gene_name,
    TRUE ~ "XXXXXXXXXXXXXXXXXXXXXXXXXXXX"
  )) %>%
  mutate(new_name = make.names(new_name))

    colnames(rnadat)<- rowdat$new_name
    
    rnadat<-data.table(rnadat)

    # genename <- gsub("\\-", ".", rnadat$genename)
    # 
    # rnadat <- as.data.table(t(rnadat[, 2:dim(rnadat)[2]]))
    # 
    # colnames(rnadat) <- genename
    
    rnadat$rna_barcode <- rna_barcodes
    
    rnadat$meta.barcode <- rna_barcodes
    
    rnadat$portion.barcode <- substr(rnadat$rna_barcode, 1, 19)

    
    rnadat$patient <- substr(rnadat$rna_barcode, 1, 12)
    
    # rnadat$sample.barcode <- substr(rnadat$rna_barcode, 1, 19)
    
    
    
    
    
    # Prepare miRNAseq data
    
    message("Preparing miRNAseq data")
    
    mir <- readRDS(paste0("miRNAseq_RDS/", i, "_miRisoform.rds" ))  
    
    mir_names <- gsub("\\-", ".", mir$mir_name)  
    
    mir[is.na(mir)] <- 0
    
    mir <- cpm(mir[, 2:dim(mir)[2]], prior.count = 0, log = F)
    
    mir <- log(mir+1, base = 10)
    
    mir_barcodes <- colnames(mir)[2:dim(mir)[2]]
    
    mir <- as.data.table(t(mir[, 2:dim(mir)[2]]))
    
    colnames(mir) <- mir_names
    
    mir$mir_barcode <- mir_barcodes
    
    # mir$meta.barcode <- mir_barcodes

    # mir$sample.barcode <- substr(mir$mir_barcode, 1, 22)
    
    # mir$patient <- substr(mir$mir_barcode, 1, 12)
  
    mir$portion.barcode <- substr(mir$mir_barcode, 1, 19)

    
    mir <- mir[, -"nonmature_mir"]
    
    
    message("Creating merged data frame")
    
    # Create merged dataframe
    
    #'[sdgfsdgfdfgdfgyjuj]
    
    rnadat<-rnadat[!duplicated(substr(rnadat$rna_barcode,1,19)),]
    
    mir<-mir[!duplicated(substr(mir$mir_barcode,1,19)),]
    
    merged <- merge(rnadat, coldat, by="meta.barcode")
    
    merged<- merge(merged, mir,all.x=TRUE, by="portion.barcode")
    
    if(!i %in% c("DLBC", "LAML", "MESO","OV", "TGCT", "THYM")) merged <- merge(merged, subtype, all.x=T, by.x="patient", by.y="meta.patient")
    
    merged <- merged[, !grepl("\\.y$", colnames(merged)), with=F]
    
    colnames(merged) <- gsub("\\.x$", "", colnames(merged))
    
    colnames(merged) <- gsub(" |\\-", ".", colnames(merged))
    
    dupcols <- duplicated(tolower(colnames(merged)))
    
    merged <- merged[, !dupcols, with=F]
    
    message("Saving RDS file")
    
    setcolorder(merged, setdiff(names(merged), c("patient", "sample_barcode")))
    
    colnames(merged) <- gsub("^patient$", "meta.patient", colnames(merged))
    colnames(merged) <- gsub("^sample_barcode$", "meta.sample_barcode", colnames(merged))
    
    merged <- merged[, !duplicated(colnames(merged)), with=F] 
    
    saveRDS(merged, paste0("projects/", i, ".rds"))
    
}

```

# Add PanImmune classification (Thorsson, Cell, 2018)

Supplementary table-S1 contains information for 11080 patients.

From Methods section of the article:

> For the molecular data matrices above, a single representative aliquot was selected per participant for cases where more than one aliquot was available, as follows. When data on more than one tumor sample was available, a choice of primary tumor sample was favored, and in remaining cases metastatic were selected over ''additional metastatic.'' For gene expression, a handful of cases were not resolved by these rules and the following aliquots were adopted based on BCR annotations:

> TCGA-23-1023: TCGA-23-1023-01A-02R-1564-13 TCGA-06-0156-01:TCGA-06-0156-01A-02R-1849-01 TCGA-06-0211-01:TCGA-06-0211-01B-01R-1849-01 TCGA-21-1076-01: TCGA-21-1076-01A-01R-0692-07

```{r}


library(data.table)

# Prepare pan immune dataset (S1 table)
panimm <- readxl::read_xlsx("Thorsson_Cell_S1.xlsx", na = "NA")
colnames(panimm)[1] <- "patient"
colnames(panimm) <- gsub("\\-| ", ".", colnames(panimm))
colnames(panimm) <- paste0("meta.", colnames(panimm))
colnames(panimm) <- tolower(colnames(panimm))
panimm <- as.data.table(panimm)


files <- gsub(".rds", "", list.files("projects/"))

for(i in files){
  
  message(paste("Processing", i))

dat <- readRDS(paste0("projects/", i, ".rds"))


# getwd()


cols <- colnames(panimm)[colnames(panimm) != "meta.patient"]

# Priority for adding panimmune data to merged dataset 
# https://gdc.cancer.gov/resources-tcga-users/tcga-code-tables/sample-type-codes

remap <- c(TP = 1,  # primary solid tumor
           TB = 2,   # primary blood derived cancer from blood
           TBM = 3, # primary blood derived cancer from BM
           TM = 4, # metastatic solid tumor
           TR = 5, # recurrent solid tumor
           TRB = 6, # recurrent blood derived cancer from blood
           TRBM = 7, # recurrent blood derived cancer from BM
           TAP = 8, # additional new primary
           TAM = 9 # additional new metastatic
           )

# below code works but it is slow. Try speeding up below by working on DT subset
# dat[, meta.remapped := remap[meta.shortLetterCode]]
# setkey(dat, meta.patient, meta.remapped)
# dat[rowid(meta.patient)==1L, (cols) := panimm[.SD, on=.(meta.patient), mget(cols)]]
# dat <- dat[is.na(meta.remapped), (cols) := NA]




dat[, meta.remapped := remap[meta.shortLetterCode]]

setkey(dat, meta.patient, meta.remapped)


dat_trim <- subset(dat, select= c("meta.patient", "meta.remapped"))

dat_trim[rowid(meta.patient)==1L, (cols) := panimm[.SD, on=.(meta.patient), mget(cols)]]

dat_trim <- dat_trim[is.na(meta.remapped), (cols) := NA]

dat_trim <- dat_trim[, -c("meta.patient", "meta.remapped")]

dat <- cbind(dat, dat_trim)

dat <- dat[, -c("meta.remapped")]

saveRDS(dat, paste0("projects/", i, ".rds"))

}

```

# Recode NA factor levels

```{r}

library(data.table)


files <- gsub(".rds", "", list.files("projects/"))

for(i in files){
  
  message(paste("Processing", i))

dat <- readRDS(paste0("projects/", i, ".rds"))

dat <- as.data.table(lapply(dat, function(x){
  
  if(class(x) == "factor")forcats::fct_explicit_na(x, "NA") else x 
  
})
)

saveRDS(dat, paste0("projects/", i, ".rds"))

}


```

# Recode character columns as factors

```{r}

library(data.table)


files <- gsub(".rds", "", list.files("projects/"))

for(i in files){
  
  message(paste("Processing", i))

dat <- readRDS(paste0("projects/", i, ".rds"))

dat <- as.data.table(lapply(dat, function(x){
  
  if(class(x) == "character") {
    x <- factor(x)
    forcats::fct_explicit_na(x, "NA")} else x 
  
})
)

saveRDS(dat, paste0("projects/", i, ".rds"))

}


```

# Manually edit certain data types

```{r}

skcm <- readRDS("projects/SKCM.rds")

grep("purity", colnames(skcm), ignore.case = T, value = T)

class(skcm$meta.subtype_PURITY..ABSOLUTE.)
class(skcm$meta.PURITY..ABSOLUTE.)


skcm$meta.subtype_PURITY..ABSOLUTE.
skcm$meta.PURITY..ABSOLUTE.  #more data present


skcm$meta.purity <- as.character(skcm$meta.PURITY..ABSOLUTE.) 
skcm$meta.purity <- gsub(",",".",skcm$meta.purity)


skcm$meta.purity <- as.numeric(skcm$meta.purity)

saveRDS(skcm, "projects/SKCM.rds")

```


# Zero Count Filtration Step


```{r}

# This part is separated from this Rmd file because of the conflict issues based on specific packages, please check the "Zero_Count_Filtration_Step" Rmd file.

```

